### 2014-08-22 text/Chapter04.tex
Drop recursion section (Chapter 4)
Calculus and semantics work (Chapter 5)

@@

Wadler and Thiemann famously show a correspondence between effect systems to monads 
\cite{monads-effects-marriage}, relating effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ 
to monadic computations $M^\sigma \tau_1 \rightarrow \tau_2$. In this chapter, we show a similar
correspondence between \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry 
of $\lambda$-calculus, this is not a simple mechanical dualization.

@@

@@

\subsection{Examples of flat coeffects}

@@

@@

@@

@@

\subsection{Language with recursion}

\begin{equation*}
\tyrule{loop}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \ident{bool} &
    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \ident{unit} }
  { \coctx{\Gamma}{\cclrd{r} \;\cseq\; (\cclrd{s} \,\cseq\, \cclrd{r})^{\cclrd{*}} } \vdash \kvd{while}~e_1~\kvd{do}~e_2 : \ident{unit} }
\end{equation*}

\begin{equation*}
\tyrule{rec}
  { \coctx{\Gamma, x:\tau_1 \xrightarrow{\cclrd{\phi}} \tau_2}{\cclrd{s}} \vdash e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
  { \coctx{\Gamma}{\cclrd{s^{\ast}}} \vdash \kvd{fix}~x. e : \tau_1 \xrightarrow{\ident{fix}~\cclrd{\phi}.\cclrd{r}} \tau_2 }
\end{equation*}

\section{Related work}

\subsection{What can be monad}
\label{sec:flat-related-monads}

The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
semantic model (monads vs. comonads) is very different, we can consider extending the two to 
obtain equivalent syntactic rules. To allow splitting of implicit parameters in lambda abstraction, 
the reader monad needs an operation that eagerly performs some effects of a function: 
$(\tau_1 \rightarrow \mtyp{r \oplus s}{\tau_2}) \rightarrow \mtyp{r}{(\tau_1 \rightarrow \mtyp{s}{\tau_2})}$.
To obtain a pure lambda abstraction
for coeffects, we need to restrict the $\cpar_{r, s}$ 
operation of indexed comonads, so that the first parameter is annotated with $\cunit$ (meaning
no effects): $\C^{\cunit} A \times \C^r B \rightarrow \C^{r}(A \times B)$.

>>>>

% ==================================================================================================
%                                                                                      
%     ###           #                       #                 #       #                 
%      #            #                       #                 #                         
%      #    # ##   ####   # ##    ###    ## #  #   #   ###   ####    ##     ###   # ##  
%      #    ##  #   #     ##  #  #   #  #  ##  #   #  #   #   #       #    #   #  ##  # 
%      #    #   #   #     #      #   #  #   #  #   #  #       #       #    #   #  #   # 
%      #    #   #   #  #  #      #   #  #  ##  #  ##  #   #   #  #    #    #   #  #   # 
%     ###   #   #    ##   #       ###    ## #   ## #   ###     ##    ###    ###   #   # 
%                                                                                     
% ==================================================================================================

@@

Wadler and Thiemann famously show a correspondence between effect systems to monads 
\cite{monads-effects-marriage}, relating effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ 
to monadic computations $\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar
correspondence between \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry 
of $\lambda$-calculus, this is not a simple mechanical dualization.

@@

% ==================================================================================================
%                                                            
%       ###           ##                   ##                 
%      #   #           #                    #                 
%      #       ###     #     ###   #   #    #    #   #   ###  
%      #          #    #    #   #  #   #    #    #   #  #     
%      #       ####    #    #      #   #    #    #   #   ###  
%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
%       ###    ####   ###    ###    ## #   ###    ## #  ####  
%                                                           
% ==================================================================================================

@@

\subsection{Examples of flat coeffects}
\label{sec:flat-calculus-examples}

@@

% ==================================================================================================
%                                              
%      #####  #                                 
%        #    #                                 
%        #    # ##    ###    ###   # ##   #   # 
%        #    ##  #  #   #  #   #  ##  #  #   # 
%        #    #   #  #####  #   #  #      #  ## 
%        #    #   #  #      #   #  #       ## # 
%        #    #   #   ###    ###   #          # 
%                                         #   # 
%                                          ###                                                                         
% ==================================================================================================

@@

% ==================================================================================================
%                                                                   
%       ###                  #                    #       #          
%      #   #                 #                    #                  
%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
%          #  #  ##  #   #   #      ####  #       #       #    #     
%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
%             #   #                                                  
%              ###                                                   
%
% ==================================================================================================

@@

% ==================================================================================================
%
%      #####          #                            #                        
%      #              #                                                     
%      #      #   #  ####    ###   # ##    ###    ##     ###   # ##    ###  
%      ####    # #    #     #   #  ##  #  #        #    #   #  ##  #  #     
%      #        #     #     #####  #   #   ###     #    #   #  #   #   ###  
%      #       # #    #  #  #      #   #      #    #    #   #  #   #      # 
%      #####  #   #    ##    ###   #   #  ####    ###    ###   #   #  ####  
%
% ==================================================================================================

@@

% ==================================================================================================
%                                                 
%      ####           ##            #                # 
%      #   #           #            #                # 
%      #   #   ###     #     ###   ####    ###    ## # 
%      ####   #   #    #        #   #     #   #  #  ## 
%      # #    #####    #     ####   #     #####  #   # 
%      #  #   #        #    #   #   #  #  #      #  ## 
%      #   #   ###    ###    ####    ##    ###    ## # 
%
% ==================================================================================================

\section{Related work}
\label{sec:flat-related-monads}

Most of the related work has been already discussed in Chapter~\ref{ch:pathways} or throughout
the chapter. In this section, we look in more details at the relationship between coeffects,
comonads and monads.

\subsection{When coeffect is not a monad}
Coeffect systems differ from effect systems in three important ways:

\begin{itemize}
\item Semantically, coeffects capture very different notions of computation. As demonstrated in 
  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
  computation, many of which cannot be captured by a monad (\eg~liveness or data-flow).

\item Syntactically, coeffect calculi use a richer algebraic structure with point-wise composition,
  sequential composition and context merging $(\cpar, \cseq, \czip)$ while most effect systems
  only use a single operation for sequential composition (monadic bind).

\item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In 
  coeffect systems, the context requirements of the body can be split between declaration-site
  and call-site, while monadic effect systems delay all effects.
\end{itemize}

\noindent
Despite the differences, our implicit parameters example can be also represented by a monad.
Semantically, the \emph{reader} monad is equivalent to the \emph{product} comonad. Syntactically,
we use the $\cup$ operation for all three operations of the coeffect algebra. However, the last
point requires us to extend monadic lambda abstraction.

% -------------------------------------------------------------------------------------------------

\subsection{When coeffect is a monad}
As discussed in \ref{sec:applications-flat-impl} ...

Product comonad

$C_{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma)$

Reader monad

$M_{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau$

\begin{equation*}
\begin{array}{l}
 \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 = \\
 (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
 \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) \\
 \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \\
\end{array}
\end{equation*}

The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
semantic model (monads vs. comonads) is very different, we can consider extending the two to 
obtain equivalent syntactic rules. To allow splitting of implicit parameters in lambda abstraction, 
the reader monad needs an operation that eagerly performs some effects of a function:

\begin{equation*}
(\tau_1 \rightarrow \mtyp{\cclrd{r} \cseq \cclrd{s}}{\tau_2}) \rightarrow 
  \mtyp{\cclrd{r}}{(\tau_1 \rightarrow \mtyp{\cclrd{s}}{\tau_2})}
\end{equation*}

To obtain a pure lambda abstraction
for coeffects, we need to restrict the $\cpar_{r, s}$ 
operation of indexed comonads, so that the first parameter is annotated with $\cunit$ (meaning
no effects): $\C^{\cunit} A \times \C^r B \rightarrow \C^{r}(A \times B)$.

\begin{equation*}
\begin{array}{rcl}
 \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
 \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
\end{array}
\end{equation*}
% ==================================================================================================
